---
title: "01-simulate"
format: html
bibliography: references.bib
---

## Simulate data

In this file, I simulate data to test the Stan model which will be used to model the actual data from two pre-existing datasets [@coretta2018l and @coretta2019g].

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
theme_set(theme_light())
library(faux)
library(cmdstanr)
library(brms)
library(posterior)
library(MASS)
```

## Prepare data

This section simulates vowel duration data.
The following code sets up the following contextual variables:

-   `P_L` is the number of participants (15) per language.

-   `L` is the number of languages (3, in the real data these are Italian, Polish and Mancunian English).

-   `V` is the number of voicing contexts of the consonant following the vowel (2, voiceless and voiced).

-   `S` is the number of syllabic contexts (2, monosyllabic and disyllabic words).

-   `P` is the total number of participants, i.e.
    `P_L * L` participants per language times number of languages.

-   `P_id` is a participant ID.

The following parameters define the regression coefficients and the multilevel hyperparameters:

-   `int_1, int_2, int_3` are the intercepts of vowel duration for each language for disyllabic words and voiceless context (we are using logged durations here).

-   `voi_1, voi_2, voi_3` are the coefficients for the difference in duration between voiced and voiceless contexts in the three languages, for disyllabic words.

-   `rat` is the effect of speech rate (same for all languages).

-   `syl` is the effect of monosyllabic words and `voi_syl` is the interaction effect of voiced consonants in monosyllabic words.

-   The `sigma_*` multilevel hyperparameters define the multilevel SDs for participant for each of the regression coefficients.
    `rho` is the correlation between the multilevel hyperparameters.

```{r}
#| label: pars

# Contextual vars
P_L <- 15
L <- 3
V <- 2
S <- 2
P <- P_L * L
P_id <- 1:P

# Regression coefficients
int_1 <- 4.5
int_2 <- 4.5
int_3 <- 4.0
voi_1 <- 0.3
voi_2 <- 0.15
voi_3 <- 0.15

rat <- -0.3
syl <- 0.1
voi_syl <- 0.1

# Multilevel hyperparameters
sigma_int <- 0.15
sigma_voi <- 0.1
sigma_rat <- 0.005
sigma_syl <- 0.005
sigma_voi_syl <- 0.01
rho <- 0.4
```

With the regression coefficients and hyperparameters we can simulate participant specific effects in `r_part`.
`Sigma` is a variance-covariance matrix needed to simulate from a multivariate Gaussian distribution with `mvnorm()`.
Since the coefficients of intercept `int_*` and voicing `voi_*` are language specific, we set their mean to 0 in `Mu`, so `r_part` has participant deviations from their language specific effect.

```{r}
#| label: part-effects

# int/voi mean = 0 because it is adjusted by lang specific effects below
Mu <- c(0, 0, rat, syl, voi_syl)
sigmas <- c(sigma_int, sigma_voi, sigma_rat, sigma_syl, sigma_voi_syl)

Rho <- matrix(rho, nrow = 5, ncol = 5)
diag(Rho) <- 1
Sigma <- diag(sigmas) %*% Rho %*% diag(sigmas)

set.seed(8)
r_part <- mvrnorm(P, Mu, Sigma)
```

Now we can adjust the intercept and voicing effects for each participant depending on their language.
We assign the first `P_L` participants to language 1, the second to language 2 and the third to language 3.
We add a language index `lang` to `r_part`.

```{r}
#| label: lang

# Adjust int/voi per language
r_part[,1] <- r_part[,1] +
  c(rep(int_1, P_L), rep(int_2, P_L), rep(int_3, P_L))
r_part[,2] <- r_part[,2] +
  c(rep(voi_1, P_L), rep(voi_2, P_L), rep(voi_3, P_L))

# Language index
lang <- rep(1:L, each = P_L)
r_part <- cbind(r_part, lang)
```

Now we set up the columns for the data.
We also simulate a different baseline speech rate for each participant and simulate small rate deviations for each repetition.

```{r}
#| label: cols

N_P <- 5
N <- N_P * P * V * S
voicing <- rep(0:1, N/2)
syllables <- rep(rep(0:1, each = 2), N/4)
participant <- rep(P_id, each = N_P * V * S)
language <- rep(lang, each = N_P * V * S)

set.seed(7)
rate_P <- rnorm(P, sd = 0.2)
rate <- rnorm(N, mean = rep(rate_P, each = N_P * V * S))

```

We can now derive the mean of vowel duration `mu` depending on the predictors.
Finally we bind everything into a tibble in `sim_data`.

```{r}
#| label: sim-data

mu <- r_part[participant, 1] + r_part[participant, 2] * voicing +
  r_part[participant, 3] * rate +
  r_part[participant, 4] * syllables +
  r_part[participant, 5] * voicing * syllables
sigma <- 0.05

set.seed(8)
y <- rnorm(N, mu, sigma)

sim_data <- tibble(
  y, participant, voicing, rate, language, syllables
)
```

### Plot

```{r}
#| label: rate-plot

sim_data |> 
  ggplot(aes(rate, y)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm")
```

```{r}
#| label: voi-lang-syl-plot

sim_data |> 
  ggplot(aes(as.character(syllables), y, colour = as.character(voicing))) +
  geom_jitter(position = position_jitterdodge(dodge.width = 0.75), alpha = 0.75) +
  facet_grid(cols = vars(language)) +
  labs(colour = "Voicing", x = "Number of syllables", y = "Vowel duration (logged ms)") +
  scale_color_brewer(type = "qual")
```

The real data we will analyse in `02-modelling.qmd` has mono and disyllabic words in English but only disyllabic words in Italian and Polish.
We remove the missing conditions in the simulated data to test the bespoke Stan model.

```{r}
#| label: sim-data-simp

sim_data <- sim_data |> 
  filter(language == 1 | (language == 2 & syllables == 0) | (language == 3 & syllables == 0))
```

```{r}
#| label: sim-data-plot

sim_data |> 
  ggplot(aes(as.character(syllables), exp(y), colour = as.character(voicing))) +
  geom_jitter(position = position_jitterdodge(dodge.width = 1), alpha = 0.75) +
  facet_grid(cols = vars(language)) +
  labs(colour = "Voicing", x = "Number of syllables", y = "Vowel duration (logged ms)") +
  scale_color_brewer(type = "qual")
```

## Model with Stan

First we prepare the data for Stan (it needs to be a list with variables that match the model).
You can find the Stan model in `scripts/bm-1.stan`.

```{r}
#| label: bm-sim-dat

bm_sim_dat <- list(
  N = nrow(sim_data),
  P = length(unique(sim_data$participant)),
  L = 3,
  y = sim_data$y,
  voicing = sim_data$voicing,
  rate = sim_data$rate,
  # di = 0 and mono = 1
  syllables = sim_data$syllables,
  participant = sim_data$participant,
  language = sim_data$language
)
```

The following code compiles the Stan file into an executable model.
The executable is saved as `scripts/bm-1` so you might not need to recompile it (running the code will just reread the compiled model).
The model was compiled with CmdStan 2.38.0 (if you have a different version you might have to recompile the model).

```{r}
#| label: bm-sim-mod

bm_sim_mod <- cmdstan_model("scripts/bm-1.stan")
```

```{r}
#| label: bm-sim-fit

mod <- "data/cache/bm_sim_fit.rds"

if (file.exists(mod)) {
  bm_sim_fit <- readRDS(mod)
} else {
 bm_sim_fit <- bm_sim_mod$sample(
    data = bm_sim_dat,
    seed = 2183,
    chains = 4, parallel_chains = 4,
    # ~1-2% divergent transitions, increase adapt delta
    adapt_delta = 0.99,
    # 7 transitions hit max treedepth, increase
    max_treedepth = 12
  )
 bm_sim_fit$draws()
 saveRDS(bm_sim_fit, mod)
}
```

```{r}
#| label: bm-sim-draws
bm_sim_draws <- bm_sim_fit$draws(
  c("intercept", "b_voicing", "b_syllables_en", "b_voicing_syllables_en", "b_rate", "r_participant", "corr_participant")
) |> as_draws_df()
```

The parameters values are recovered quite well.

```{r}
#| label: bm-sim-summary
bm_sim_fit$summary(variables = c("intercept", "b_voicing", "b_syllables_en", "b_voicing_syllables_en", "b_rate")) |> 
  mutate(
    across(where(is.numeric), ~round(.x, 2))
  )
```
